Behind Bars and Beyond: Investigating Socioeconomic Drivers of Incarceration Rates in the United States

### 2024-04-17 paper/paper.qmd
Updated data, finished data section, and created model.
Removed an unneeded variable from the data, finalized data section in quarto, and updated model

@@

# Results {#sec-results}

>>>>

The finalized cleaned dataset used in the analysis that will follow was created by merging the aforementioned 5 datasets into one. In particular, only countries that were present in all 5 datasets are present in the cleaned dataset, bringing the total number of nations included in this study to 134. Further, the cleaned analysis dataset was filtered to only contain the following information from the 5 datasets: a country's name, its incarceration rate (per 100k population), its national poverty rate, its unemployment rate, its violent crime rate, and its educational ranking category. Note that a nation's violent crime rate and educational ranking category were not explicit data that were provided as part of the 5 raw datasets. Instead, these were manually constructed and appended to the final dataset during the cleaning process.

In this study, a nation's violent crime rate is defined to be the sum of all the crime rates (per 100k population) available through the violent crime dataset in order to have one final value that represents the extent of violent crime in a nation. It is important to note, however, that various nations had missing values for some of the violent crime statistics, and as a result, these values were simply treated as 0 for the sake of the sum. As for the educational ranking category, as opposed to treating educational ranking as a continuous variable in the analysis that will follow, it was instead decided to create 10 categories representing the 10 deciles of educational ranks present in the final dataset, based on the ranks of the 134 countries present, essentially splitting the countries into 10 equally sized categories by educational ranking. This was done in order to have a categorical variable whose coefficient will be more easily interpreted in the model present in @sec-model, as interpreting a 1 unit increase in educational ranking can be quite arbitrary, as moving up or down one rank educationally may be more subjective. @tbl-cleaned-data

```{r}
#| label: tbl-cleaned-data
#| tbl-cap: Top 5 countries by incarceration rate in the finalized cleaned dataset
#| echo: false

cleaned_data |>
  arrange(desc(incarceration_rate)) |>
  head(5) |>
  kable(
        col.names = c("Country", "Incarceration Rate", "Poverty Rate", "Unemployment Rate", "Educational Ranking Category", "Violent Crime Rate")
  )
```

To gain a better understanding of the distribution of incarceration rates as a function of the independent variables, the figures below each showcase all observations, placing incarceration rates on the y-axis and independent variable on the x-axis. Moreover, we have chosen to highlight 5 particular nations in each of the figures, representing nations across the entire range of incarceration rates present in the dataset. In particular, we highlight El Salvador who has the highest incarceration rate of 1086, the United States with the 3rd highest incarceration rate of 531, Mexico with the 51st highest incarceration rate of 174, Iceland with the 5th lowest incarceration rate of 36, and Guinea-Bissau with the lowest incarceration rate of 31. Do note that these rankings are relative to the finalized dataset as opposed to the entire list of countries and territories found in the raw incarceration rate dataset.

@fig-poverty-incarceration showcases all observations by poverty rate, highlighting the aforementioned countries as well. Similarly, @fig-unemployment-incarceration showcases all observations by unemployment rate and @fig-violent-crime-incarceration by violent crime rates. Lastly, @fig-education-incarceration showcases the mean incarceration rate for each of the 10 educational ranking categories. Note that Iceland is in the "1-19" category, the United States is in the "20-34" category, Mexico is in the "35-52" category, El Salvador is in the "111-125" category, and Guinea-Bissau is in the "176-198" category.

```{r}
#| label: fig-poverty-incarceration
#| fig-cap: Scatter plot of incarceration rates by poverty rates
#| echo: false
#| fig-pos: H

ggplot(cleaned_data, aes(x = poverty_rate, y = incarceration_rate)) +
  geom_point(color = "darkgrey") +
  geom_point(data = cleaned_data |> filter(country %in% c("United States", "El Salvador", "Guinea-Bissau", "Iceland", "Mexico")),
             aes(color = country), size = 1.5) +  # Highlight specific countries
  theme_classic() +
  labs(x = "Poverty Rate", y = "Incarceration Rate")
```

```{r}
#| label: fig-unemployment-incarceration
#| fig-cap: Scatter plot of incarceration rates by unemployment rates
#| echo: false
#| fig-pos: H

ggplot(cleaned_data, aes(x = unemployment_rate, y = incarceration_rate)) +
  geom_point(color = "darkgrey") +
  geom_point(data = cleaned_data |> filter(country %in% c("United States", "El Salvador", "Guinea-Bissau", "Iceland", "Mexico")),
             aes(color = country), size = 1.5) +  # Highlight specific countries
  theme_classic() +
  labs(x = "Unemployment Rate", y = "Incarceration Rate")
```

```{r}
#| label: fig-violent-crime-incarceration
#| fig-cap: Scatter plot of incarceration rates by violent crime rates
#| echo: false
#| fig-pos: H

ggplot(cleaned_data, aes(x = violent_crime_rate, y = incarceration_rate)) +
  geom_point(color = "darkgrey") +
  geom_point(data = cleaned_data |> filter(country %in% c("United States", "El Salvador", "Guinea-Bissau", "Iceland", "Mexico")),
             aes(color = country), size = 1.5) +  # Highlight specific countries
  theme_classic() +
  labs(x = "Violent Crime Rate", y = "Incarceration Rate")
```

```{r}
#| label: fig-education-incarceration
#| fig-cap: Bar chart showcasing mean incarceration rate for each educational ranking category
#| echo: false
#| fig-pos: H

# Define the levels in the desired order
education_levels <- c("1-19", "20-34", "35-52", "53-71", "72-92", 
                      "93-110", "111-125", "126-147", "148-175", "176-198")

# Convert education_category to a factor with specified levels
cleaned_data$education_category <- factor(cleaned_data$education_category, levels = education_levels)

# Recalculate the mean incarceration rate for each educational category
mean_incarceration <- cleaned_data |>
  group_by(education_category) |>
  summarize(mean_incarceration_rate = mean(incarceration_rate, na.rm = TRUE))

# Create the bar chart
ggplot(mean_incarceration, aes(x = education_category, y = mean_incarceration_rate)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  theme_classic() +
  labs(x = "Educational Ranking Category", y = "Mean Incarceration Rate") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability

```

To gain a better understanding of the underlying data, @tbl-summary-stats provides information on the overall mean and standard deviation of each of the variables present in the cleaned dataset.

```{r}
#| label: tbl-summary-stats
#| tbl-cap: Mean and standard deviation for each variable
#| echo: false
#| fig-pos: H

# Calculate summary statistics
mean_incarceration <- round(mean(cleaned_data$incarceration_rate, na.rm = TRUE), 2)
sd_incarceration <- round(sd(cleaned_data$incarceration_rate, na.rm = TRUE), 2)
mean_poverty <- round(mean(cleaned_data$poverty_rate, na.rm = TRUE), 2)
sd_poverty <- round(sd(cleaned_data$poverty_rate, na.rm = TRUE), 2)
mean_unemployment <- round(mean(cleaned_data$unemployment_rate, na.rm = TRUE), 2)
sd_unemployment <- round(sd(cleaned_data$unemployment_rate, na.rm = TRUE), 2)
mean_violent <- round(mean(cleaned_data$violent_crime_rate, na.rm = TRUE), 2)
sd_violent <- round(sd(cleaned_data$violent_crime_rate, na.rm = TRUE), 2)

# Create a data frame manually
data <- data.frame(
  Variable = c("Incarceration Rate", "Poverty Rate", "Unemployment Rate", "Violent Crime Rate"),
  Mean = c(mean_incarceration, mean_poverty, mean_unemployment, mean_violent),
  SD = c(sd_incarceration, sd_poverty, sd_unemployment, sd_violent)
)

# Print the data frame as a table using kable
kable(data, col.names = c("Variable", "Mean", "SD"))
```

@@

# Results {#sec-results}
Make note that poorer nations could also be incapable of housing more individuals in jails (use kenya as case study. 284% jail capacity)
